Staged Abstract Interpreters

### 2019-09-04 paper/20_prelim.tex
revision

Similar to Haskell's @do@-notation, Scala provides special syntactic support for
monadic operations through @for@-comprehension.
For example, an object of @List[A]@ is an instance of @List@ monad, where @A@ is the element type. 
Then to compute the Cartesian product of two lists of numbers, we can use Scala's
@for@-comprehension syntax.
\begin{lstlisting}
  val xs = List(1, 2); val ys = List(3, 4)
  for { x <- xs; y <- ys } yield (x, y)   // List((1,3), (1,4), (2,3), (2,4))
\end{lstlisting}

The Scala compiler will translate the above @for@-comprehension expression into
an equivalent one using @flatMap@ and @map@ ~\cite{scala_spec}. The last binding
in the @for@-comprehension is translated into a @map@, where the expression of
@yield@ becomes the body expression of that @map@ application. The foregoing
bindings in the comprehension are all translated into calls of @flatMap@.
\begin{lstlisting}
  xs.flatMap { case x => ys.map { case y => (x, y) } }
\end{lstlisting}

Note that here the monadic object @List[_]@ encapsulates the data
internally.  Therefore it only exposes the simplified version of
@flatMap@, where the monad @M[A]@ is not introduced as a function
argument. The trait @MonadOps@ (Figure \ref{fig:monadops}) defines the
simplified version of monadic operations that are necessary for
@for@-comprehension. The conversion between @Monad@ and @MonadOps@
can be done by using the implicit design pattern.
In the rest of the paper, we use Scala's @for@-comprehension syntax
and a few of monads and monad transformers such as @ReaderT@,
@StateT@, and @ListT/SetT@ to write our interpreters. The implementation
of monads and monad transformers essentially borrows the ground-truth
from Haskell.

@@

Monad transformers are type constructors of kind @(* -> *) -> (* -> *)@, which
take a monad as argument and produces another monad. By using monad
transformers, we can combine multiple monads into a single one. Constructing
extensible interpreters using monad transformers was first explored by
\citet{DBLP:conf/popl/LiangHJ95}, and later has been applied to abstract
interpreters \cite{Sergey:2013:MAI:2491956.2491979,
  DBLP:journals/pacmpl/DaraisLNH17, Darais:2015:GTM:2814270.2814308}.
In this section, we present the generic interface in the style of big-step
definitional interpreter. The key idea is to keep the binding-time type and
returned monadic type abstract, so that they can be instantiated differently.
We also need to abstract the primitive operations on those types.

\paragraph{Basic Types} We start with some basic type definitions used in the
interpreter. The identifiers in the program are represented by
strings. To represent states, two required components for the
interpreter are environments @Env@ and stores @Store@. @Env@ maps from
identifiers to addresses, and @Store@ maps from addresses to values,
respectively. The @Env@ captures bound variables in scope, and @Store@
models a persistent heap through the program run-time.  At this moment,
the domains of addresses and values are still abstract, i.e., they are
declared as abstract types.
\begin{lstlisting}
  trait Semantics {
    type Ident = String; type Addr; type Value
    type Env = Map[Ident, Addr]; type Store = Map[Addr, Value]
    type R[_] // Binding-time as a higher-kinded type
    ... // The definitions in the rest of this section are enclosed in trait Semantics.
  }
\end{lstlisting}

\paragraph{Binding-time Abstraction} As mentioned before, the binding-time is
declared as a higher-kinded type @R[_]@. We are free to use @R@ to annotate
other data-types in the interpreter; it is also injected into the monadic
type @MonadOps@.
If we simply instantiate @R@ an identity type (i.e., @type R[T] = T@),
then the generic interpreter is a standard definitional interpreter
that will execute the program.  In Section \ref{stagedinterp}, we will
instantiate @R@ using LMS's built-in next-stage type annotation @Rep@,
which makes the interpreter act as a compiler.

\paragraph{Monadic Operations} We define the return type of the interpreter as
@Ans@, which is a monad type @AnsM[_]@ wrapping the type @Value@. 
As mentioned in Section ~\ref{monadscala}, to use the @for@-comprehension
syntax, certain operations have to be added on the type @AnsM@. Here, we use a
structural type @MonadOps@ to require that type @AnsM@ at least implements @map@ and
@flatMap@. It is worth noting that @MonadOps@ takes another type parameter
@R[_]@ as binding-time that annotates data; accordingly the type @R@
defined in the outer scope is passed to @MonadOps@. Inside of @MonadOps@,
@R[_]@ wraps the data types @A@ and @B@ that are encapsulated by the
monad, but not the monad type @M@ itself. When acting as compilers, we
will also replace the monads to the ones that work on staged data values.

We also declare several operations to manipulate environments and
stores. These methods return monadic values of type @AnsM[_]@, which
may encapsulate over the environment or store, or simply a @Unit@
value for effects. For examples, @local_env@ installs a new environment
@ρ@ when evaluating the monadic value @ans@; @set_store@ takes a pair of
addresses and values (potentially staged) and updates the store accordingly.

@@

\paragraph{Primitive Operations} Next we define several primitive operations.
First, two versions of @alloc@ are declared. The first one takes a store and an
identifier and produces a fresh address of non-monadic type @R[Addr]@. Since
the freshness of the address may depend on the store, which might be a
next-stage value as indicated by its type, the type of addresses is therefore
wrapped by @R[_]@. The other @alloc@ is simply the monadic version of the addresses.
\begin{lstlisting}
  def alloc(σ: R[Store], x: Ident): R[Addr];  def alloc(x: Ident): AnsM[Addr]
\end{lstlisting}

Other primitive operations provide basic functionality for the
interpreter.  The method @num@ and @close@ deal with primitive values,
which lift literal terms (e.g., lambdas) to our value representation
(e.g., closures).  The method @get@ retrieves the value of an
identifier @x@, through the environment and store. Conditionals and
arithmetic are handled by @br0@ and @arith@, respectively. The method
@ap_clo@ is used for function applications: it takes a function value
and an argument value. Note that the @Env@, @Store@, and @Value@ are
all annotated by @R[_]@ since they can potentially be next-stage
values when acting as compilers.
\begin{lstlisting}
  def num(i: Int): Ans
  def get(σ: R[Store], ρ: R[Env], x: Ident): R[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: R[Env]): R[Value]
  def br0(test: R[Value], thn: => Ans, els: => Ans): Ans
  def arith(op: Symbol, v1: R[Value], v2: R[Value]): R[Value]
  def ap_clo(ev: Expr => Ans)(rator: R[Value], rand: R[Value]): Ans
\end{lstlisting}

@@

\paragraph{The Interpreter} Now we construct the semantics-agnostic interpreter
in monadic form, shown in Figure \ref{fig:shared_int}. The basic idea
of a generic interpretation is to traverse the abstract syntax tree
while maintaining the effects such as reader and state.  Note that the
interpreter is written in open-recursive style -- it can not refer to
itself directly, instead, @eval@ takes an additional parameter @ev@ of
type @Expr => Ans@ that refers to itself. Consequently, the method
@close@ that lifts lambda terms to closures and method @ap_clo@ that
applies functions also takes an extra @ev@, because evaluation may
happen inside of them.
To close the open-recursion, we use a fixed-point operator @fix@.
For concrete-interpretation instantiation, it works like the Y
combinator; for abstract-interpretation instantiation, it instruments
the interpreter by memoizing its input and output, which will ensure
the termination of abstract interpretation.
Finally, a top-level wrapper @run@ is declared; the return type
@Result@ depends on what kind of monad we will be using, this is also
an abstract type.
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans
  type Result; def run(e: Expr): Result
\end{lstlisting}

>>>>

Similar to Haskell's @do@-notation, Scala provides special syntactic support
for monadic operations through @for@-comprehension.  For example, an object of
@List[A]@ is a @List@ monad, where @A@ is the element type.
To compute the Cartesian product of two lists of numbers, we can use Scala's
@for@-comprehension syntax:
\begin{lstlisting}
  val xs = List(1, 2); val ys = List(3, 4)
  for { x <- xs; y <- ys } yield (x, y)   // List((1,3), (1,4), (2,3), (2,4))
\end{lstlisting}

The Scala compiler will translate the above @for@-comprehension expression into
an equivalent one using @flatMap@ and @map@ ~\cite{scala_spec}. The last binding
in the @for@-comprehension is translated into a @map@, where the expression of
@yield@ becomes the body expression inside of that @map@ operation. The foregoing
bindings in the comprehension are all translated into calls of @flatMap@:
\begin{lstlisting}
  xs.flatMap { case x => ys.map { case y => (x, y) } }
\end{lstlisting}

Note that here the monadic object @List[_]@ encapsulates the data internally.
Therefore it only exposes the simplified version of @flatMap@, where the monad
@M[A]@ is not introduced as a function argument. The trait @MonadOps@ (Figure
\ref{fig:monadops}) defines the simplified version of monadic operations that
are necessary for @for@-comprehension. The conversion between @Monad@ and
@MonadOps@ can be done by using implicit design pattern.
In the rest of the paper, we use Scala's @for@-comprehension syntax and a few
of monads and monad transformers such as @ReaderT@, @StateT@, and @SetT@ to
write our interpreters.
Monad transformers are type constructors of kind @(* -> *) -> (* -> *)@, which
take a monad as argument and produces another monad. By using monad
transformers, we can combine multiple monads into a single one.  The
implementation of monads and monad transformers essentially borrows the
ground-truth from Haskell.

@@

In this section, we present the generic interface in the style of big-step
definitional interpreter. The key idea is to keep the binding-time type and
returned monadic type both abstract, so that they can be instantiated
differently; we also need to abstract the primitive operations on those types.
In later sections, we will instantiate the monadic type to perform concrete
interpretation \citet{DBLP:conf/popl/LiangHJ95} or abstract interpretation
\cite{Sergey:2013:MAI:2491956.2491979, DBLP:journals/pacmpl/DaraisLNH17}.

\paragraph{Basic Types} 
We start with some basic type definitions that are used in the interpreter. The
identifiers in the program are represented by strings. To represent states, two
required components for the interpreter are environments @Env@ and stores
@Store@. @Env@ maps from identifiers to addresses, and @Store@ maps from
addresses to values, respectively. The @Env@ captures bound variables in the
scope of current control, and @Store@ models a persistent heap through the
program run-time.  At this moment, the domains of addresses and values are
still abstract, i.e., they are declared as abstract types.
\begin{lstlisting}
  trait Semantics {
    type Ident = String; type Addr; type Value
    type Env = Map[Ident, Addr]; type Store = Map[Addr, Value]
    type R[_] // Binding-time as a higher-kinded type
    ... // The definitions in the rest of this section are enclosed in trait Semantics.
  }
\end{lstlisting}

\paragraph{Binding-time Abstraction}
The binding-time is declared as a higher-kinded type @R[_]@. 
Now we are able to use @R@ to annotate other data-types in the interpreter;
it is also injected into the monadic type @MonadOps@.  If we simply instantiate
@R@ an identity type (i.e., @type R[T] = T@), then the generic interpreter is a
standard definitional interpreter that will execute the program.  In Section
\ref{stagedinterp}, we will instantiate @R@ using LMS's built-in next-stage
type annotation @Rep@, which makes the interpreter act as a compiler.

\paragraph{Monadic Operations} We define the return type of the interpreter as
@Ans@, which is a monadic type @AnsM[_]@ wrapping the type @Value@. 
As mentioned in Section ~\ref{monadscala}, to use the @for@-comprehension
syntax, certain operations have to be added on the type @AnsM@. Here, we use a
structural type @MonadOps@ to require that type @AnsM@ should at least
implement @map@ and @flatMap@. It is worth noting that @MonadOps@ takes another
type parameter @R[_]@ as binding-time, i.e., @MonadOps[R, AnsM, T]@.
Inside of @MonadOps@, @R[_]@ wraps the data types @A@ and @B@ that are
encapsulated by the monad, but not the monad type @M@ itself. When acting as
compilers, we will also replace the monads to the ones that work on staged data
values.

We also declare several operations to manipulate environments and
stores. These methods return monadic values of type @AnsM[_]@, which
may parameterized over the environment or store type, or simply a @Unit@
value for effects. For examples, @local_env@ installs a new environment
@ρ@ when evaluating the monadic value @ans@; @set_store@ takes a pair of
addresses and values (potentially staged) and updates the store accordingly.

@@

\paragraph{Primitive Operations} Next we define a few primitive operations.
First, two versions of @alloc@ are declared. The first one takes a store and an
identifier and produces a fresh address of non-monadic type @R[Addr]@. Since
the freshness of the address may depend on the store, which might be a
next-stage value as indicated by its type, the type of addresses is
consequently wrapped by @R[_]@. The other @alloc@ is simply the monadic version
of the addresses, therefore can be used in monadic computations.
An auxiliary method @get@ retrieves the value of an identifier @x@, through the
environment and store. 
\begin{lstlisting}
  def alloc(σ: R[Store], x: Ident): R[Addr];  def alloc(x: Ident): AnsM[Addr]
  def get(σ: R[Store], ρ: R[Env], x: Ident): R[Value]
\end{lstlisting}

Other primitive operations in the interpreter handles language constructs.  The
method @num@ and @close@ deal with primitive values, which lift literal terms
(e.g., lambdas) to our value representation (e.g., closures).  
Conditionals and arithmetic are handled by @br0@ and @arith@, respectively. The
method @ap_clo@ is used for applying functions, which takes a function value
and an argument value. Note that the @Env@, @Store@, and @Value@ are all
annotated by @R[_]@ because they can potentially be next-stage values when the
interpreter acts as a compiler.
\begin{lstlisting}
  def num(i: Int): Ans
  def close(ev: Expr => Ans)(λ: Lam, ρ: R[Env]): R[Value]
  def br0(test: R[Value], thn: => Ans, els: => Ans): Ans
  def arith(op: Symbol, v1: R[Value], v2: R[Value]): R[Value]
  def ap_clo(ev: Expr => Ans)(rator: R[Value], rand: R[Value]): Ans
\end{lstlisting}

@@

\paragraph{The Interpreter} Now we can construct the semantics-agnostic interpreter
in monadic form, shown in Figure \ref{fig:shared_int}. The basic idea
of a generic interpretation is to traverse the abstract syntax tree
while maintaining the effects such as environment and state updates.  
Note that the interpreter is written in open-recursive style -- it cannot
refer to itself directly, instead, @eval@ takes an additional parameter @ev@ of
type @Expr => Ans@ that refers to itself. Consequently, the method @close@ that
lifts lambda terms to closures and method @ap_clo@ that applies functions also
takes an extra @ev@, because evaluations may happen inside of them.
To close the open-recursion, we use a fixed-point operator @fix@.
For concrete-interpretation instantiation, @fix@ works like the Y combinator;
for abstract-interpretation instantiation, it will instrument the interpreter
by memoizing @ev@'s inputs and outputs, which ensures the termination of
abstract interpretation.
Finally, a top-level wrapper @run@ is declared; the return type @Result@
depends on what kind of monads is being used, therefore this is also an
abstract type.
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans
  type Result; def run(e: Expr): Result
\end{lstlisting}

### 2019-09-04 paper/70_eval.tex
revision

To evaluate how the performance can be improved by our staged abstract interpreters
approach, we scale our toy interpreter to a large subset of the Scheme language
and evaluate the analyzing time against with the unstaged versions. The result
shows that the staged versions are averagely 10+ times as fast as the unstaged
version, as strong evidence of our \textit{abstraction without regret}
approach.

\paragraph{Implementation and Evaluation Environment} We implement the abstract
interpreters for control-flow analysis, i.e., a value-flow analysis that
computes which lambda terms can be called at each call-sites in functional
languages. With other suitable abstract domains, the abstract interpreter also
can be extended to other analyses. A front-end desugars Scheme to a small core
language that makes the analyzer easy to implement. We implement two common
monovariant 0-CFA-like analysis, one is equipped with store-widening (0CFA-SW
for short), the other does not (0CFA for short). The store-widened analysis is
expected to be faster than the one without.  Our prototype implementation
currently generates Scala code; the generated Scala code will be compiled and
executed on JVM. It is possible to generate C/C++ code in the future. In the
experiments, we implement several optimizations mentioned in Section
\ref{staged_ds}, specifically, the selective caching scheme and IR-level
rewriting rules (for pairs, lists, sets, and maps) that exploit
partially-static data. These optimizations are useful to reduce the size of
generated code.

@@

\paragraph{Benchmarks}
The benchmark programs we used in the experiment are collected from several
previous papers \cite{Johnson:2013:OAA:2500365.2500604, ashley:practical,
DBLP:journals/corr/abs-1102-3676} and existing artifacts
\footnote{https://github.com/ilyasergey/reachability} for control-flow
analysis.  Some of the benchmarks are small programs designed for
experiments, such as the @kcfa-worst-@$n$ series, which are intended
to challenge $k$-CFA; while some other benchmarks are from real-world
applications, for examples, the RSA public key encryption algorithm
@rsa@. In the Figure \ref{evaluation_result}, we report the number of
AST node after parsing and desugaring (excluding comments) as a proper
measurement of program size.

@@

Figure~\ref{evaluation_result} shows the evaluation result, comparing the
performance improved by staging on the two monovariant CFAs The
\textit{unstaged} and \textit{staged} columns show the median time to finish
the analysis. The column $\frac{\text{unstaged}}{\text{staged}}$ shows the
improvement. The dash - in some cells indicates timeout.  As
expected, the staged version significantly outperforms the unstaged version on
all benchmarks. We observe that simple and homogeneous value domains used in the
benchmark usually leads to better speed-ups.  For 0CFA w/o store-widening, we
observe an average speed-up of ~17x times; for 0CFA w/ store-widening, the
average speed-up is ~7x times.  The \texttt{kcfa-worst-64} benchmark has the most
speed-up among all the tests. Although it has a comparably large size of AST,
it has a quite simple structure which only involves booleans, functions and
applications. We conjecture that the simple domain and structure allow the
generated code to be efficiently optimized by the JVM. But on the corresponding
0CFA-SW, we do not observe such great speed-up. Due to the complexity of
generated code and next-stage running platform (JVM), we leave the detailed
empirical analysis of which static analysis can benefit more from
specialization as future work.

>>>>

To evaluate how performance can be improved by our staged abstract interpreters
approach, we scale our toy interpreter to a large subset of the Scheme language
and evaluate the analyzing time against with the unstaged versions. The result
shows that the staged versions are averagely 10+ times as fast as the unstaged
version, as evidence of our \textit{abstraction without regret} approach.

\paragraph{Implementation and Evaluation Environment} We implement the abstract
interpreters for control-flow analysis, i.e., a value-flow analysis that
computes which lambda terms can be called at each call-site. 
With other suitable abstract domains, the abstract interpreter
can also be extended to other analyses. 
A front-end desugars Scheme to a small core language that makes the analyzer
easy to implement. We implement two common monovariant 0-CFA-like analysis, one
is equipped with store-widening (0CFA-SW for short), the other does not (0CFA
for short). Usually, the store-widened analysis is expected to be faster than the one
without.  Our prototype implementation currently generates Scala code; the
generated Scala code will be compiled and executed on JVM. It is possible to
generate C/C++ code in the future. In the experiments, we implement several
optimizations mentioned in Section \ref{staged_ds}, specifically, the selective
caching scheme and IR-level rewriting rules (for pairs, lists, sets, and maps)
that exploit partially-static data. These optimizations are useful to reduce
the size of generated code.

@@

\paragraph{Benchmarks}
The benchmark programs we used in the experiment are collected from
previous papers \cite{Johnson:2013:OAA:2500365.2500604, ashley:practical,
DBLP:journals/corr/abs-1102-3676} and existing artifacts
\footnote{https://github.com/ilyasergey/reachability} for control-flow
analysis.  Some of the benchmarks are small programs designed for
experiments, such as the @kcfa-worst-@$n$ series, which are intended
to challenge $k$-CFA; while some other benchmarks are from real-world
applications, for examples, the RSA public key encryption algorithm
@rsa@. In Figure \ref{evaluation_result}, we report the number of
AST nodes after parsing and desugaring (excluding comments) as a proper
measurement of program size.

@@

Figure~\ref{evaluation_result} shows the evaluation result, comparing the
performance improved by staging on the two monovariant CFAs. The
\textit{unstaged} and \textit{staged} columns show the median time (excluding
pre-compilation) to finish the analysis. The column
$\frac{\text{unstaged}}{\text{staged}}$ shows the improvement. The dash @-@ in
some cells indicates timeout.  As expected, the staged version significantly
outperforms the unstaged version on all benchmarks. 
%We observe that simple value domains used in the benchmark usually leads to
%better speed-ups, for example, the @kcfa-worst@ series only uses lambdas and
%booleans.
For 0CFA w/o store-widening, we observe an average speed-up of ~17x times; for
0CFA w/ store-widening, the average speed-up is ~7x times.
\texttt{kcfa-worst-64} has the most speed-up among all the tests.
Although it has a comparably large size of AST, it has a quite simple structure
which only involves booleans, lambda terms and applications. We conjecture that
the simple domain and structure allow the generated code to be efficiently
optimized by the JVM. But on the corresponding 0CFA-SW, we do not observe such
great speed-up. Due to the complexity of generated code and next-stage running
platform (JVM), we leave the detailed empirical analysis of which static
analysis can benefit more from the specialization as future work.
